--- 
title: "Spatial and Temporal Considerations of Flower Occurance at the Northern Agricultural Research Center Near Havre Montana"
subtitle: "Prepared for Ermatinger Lochlin"
author: 
- Primary Statistician, Statistician
- Secondary Statistician, Statistician
- Director Name, Director
authors: 
  name: Will Hammond and Gifty Osei 
  affiliation: Statistical Consulting and Research Services  

secondary: 
  name: Riley Collins
  affiliation: Statistical Consulting and Research Services
director:  
  name: Andy Hoegh
  affiliation: Instructor for Statistical Consulting Class  

abstract: |
  **Author Note:**  
  
  This material is provided to communicate advice from SCRS statisticians based on our best understanding of your research needs. We encourage you to use this report in discussions with colleagues. Please do not publish any portion of this material without permission.  
  
  © Primary Statistician
  
  When you make use of our work for publications or presentations, please be sure to acknowledge the funding we receive from NIGMS using the following: “Research reported in this publication was supported by Institutional Development Awards (IDeA) from the National Institute of General Medical Sciences of the National Institutes of Health under Awards P20GM103474, U54GM115371, 5P20GM104417, and 2U54GM104944-06. The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.  
prelim: yes
date: "**`r format(Sys.time(), '%B %d, %Y')`**"
titlepage: yes
msucolor: yes
twoside: yes
spacing: 1
bibliography: [scrs-ref.bib, packages.bib]
biblio-style: apalike
link-citations: yes
colorlinks: yes
site: bookdown::bookdown_site
documentclass: article
---


```{r import, echo = FALSE}
# Some other useful knitr options:
# fig.retina: multiplicative factor for figures' resolution
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = FALSE,
               comment = NA, size = 'footnotesize', width = 80,
               warning = FALSE, show.signif.stars = FALSE,
               fig.height=5, fig.width=10)
```

```{r packages, include=FALSE}
# automatically create a bib database for R packages, list used packages here
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'readr', 'readxl', 'dplyr', 'mosaic',
  'nlme', 'yarrr', 'predictmeans','tibble', 'ggplot2', 'effects'), 'packages.bib')
```

```{r echo=F}
# load all required packages first, edit this list as required.
library(readr)
library(readxl)
library(dplyr)
library(mosaic)
library(knitr)
library(yarrr)
library(nlme)
library(predictmeans)
library(effects)
library(tibble)
suppressMessages(library(stringr))
```


# Introduction {#intro}

Montana's most important region of wheat production is colloquially referred to as the Golden Triangle(GT). The area owes its agricultural prominence to the grasslands that they were converted from. It has been estimated that 73$\%$ of the grassland that were present in the GT prior to European settlement of region have been lost to land conversion. In the case of wheat production, the balance between an insect pest of wheat (the wheat stem sawfly) and its natural enemies has been disrupted, resulting in epidemic outbreaks of wheat steam sawfly(WSS) infestation in wheat fields.The specialist parasitoids B. cephi and B. lissogaster require a WSS larvae feeding inside a grass or wheat stem as host for their offspring. They also require diets rich in carbohydrates and other nutrients compounds to achieve maximum livelihood and increase traits associated with fecundity. Because little is known about the floral resources that may be important to the parasitoids, a two year survey was conducted of an intact grassland plot(12 ha) located within the GT. During each survey(~6 per season, every two weeks) the location, species, and quantity of actively flowering plants encountered were recorded. LiDAR digital elevation model and daily weather observations to model the spatiotemporal context of our observed flowering plants were also retrieved. This model may help us to understand which species of flora warrant further investigation as supplemental resources for the specialist parasitoids of WSS based on their temporal alignment with the braconids's phenology and observed spatial distribution. Anticipating the culmination of this project will help inform future research on providing supplemental nutrition to parasitoids of WSS in the field setting. Identifying plants native to the study region that may provide the parasitoids with similar benefits to that of buck wheat and cowpea is the big goal. 



## Data Collection 

This research describes surveys of flowering plants conducted in a 12-hectare grassland on the northern Agricultural Research Center in Montana. The site has been protected from development and livestock grazing since 1916, allowing for the present flora to be similar to what may inhabit analogous areas in the absence of human influence. The surveys were conducted every two weeks between June and September of 2019 and 2020 using belt transects, and the surveys consisted of GPS location, species, and quantity of actively flowering plants. In total, 9720 point observations of actively flowering plants from 61 different species were collected.

# Data Description and Methods

This analysis utilized data from three primary sources: a list of flower species found on the plot of land, weather data for a nearby station, and data collected at numerous points on the plot. The flower species list contained 65 species, their scientific and common names, and a conservation score. Weather data for each day in 2019 and 2020 included mean air and soil temperature and precipitation received, used to calculate growing degree days. Data collected at numerous points on the plot included flower species and quantity, as well as variables such as x and y coordinates, mean elevation, aspect, slope, and roughness. Aspect was transformed into an indicator for north facing slopes.



```{r}
## load data
library(readxl)
library(stringr)
library(lubridate)
library(readr)
library(tidyverse)
library(ggridges)
points_terrain <- read_xlsx("D:/MSU/Spring 23/Stat 575/pointsAndTerrain_tbl.xlsx")
weather <- read.csv("D:/MSU/Spring 23/Stat 575/MT_Mesonet_havrenmt_daily_data.csv")
names(weather) <- c("station", "datetime", "Air.Temperature", "Precipitation", "SoilTemp0in", "SoilTemp4in","SoilTemp8in","SoilTemp20in","SoilTemp36in")
species_key <- read_xlsx("D:/MSU/Spring 23/Stat 575/SpeciesList.xlsx", sheet = 1, skip = 1)
species_key <- species_key[,c(2, 4:8)]


## Some plots
ggplot(points_terrain, aes(x = X, y = Y, color = Spp ))+geom_point(alpha = 0.5)+theme(legend.position="none")+theme_bw()


## from the weather data



## from the species data


```


# Poisson Point Process Models

An inhomogenous poisson point process model was used to glean insights to the distribution of species in terms of favorable or unfavorable terrain characteristics. The approach used was to fit an individual model for each species. [@banerjee2014hierarchical] These models are based on two main assumptions: 

1. Observed points are independent of each other. That is, the presence of one point in a region, $B \subset \mathbb{R}^2$, is not related to the presence of other points in the same region.

2. The number of points (flower observations) within a region,B, has a poisson distribution, and the expected count of points within a region is = $\int_{region}\lambda(u)du$, where $\lambda(u)$ is the intensity (average number of points per unit area) of points at the location $u \in B$.

Each model produces an intensity surface across the survey area according to log-linear poisson likelihood, where the points $u$ correspond to pixels in the survey area.

$$P(N(u)) \sim Pois(\lambda(u))$$

To estimate the intensity with respect to the terrain covariates we use a log-linear intensity function:

$$ \lambda(u)=exp(\beta_0 + \beta_1aspect_u + \beta_2slope_u + \beta_3roughness_u) $$

Here, the log-linear intensity function imposes another assumption, that intensity increases or decreases exponentially as a function of aspect, slope, and roughness. 

Aspect was defined as a 1 for northfacing areas and 0 for south facing areas. The coefficients $\beta$ represent the expected increase in log-intensity for a 1 unit increase in the corresponding variable (aspect, roughness, slope). $\beta$ with large magnitude represent strong underlying relationships with terrain covariates for a given species, while $\beta$ close to 0 indicate that the species is relatively unaffected by the terrain covariate.  

The model coefficients then represent estimates of the relationship between a species presence at a given location with observed terrain characteristics. For the context of understanding the individual terrain preferences of a species, the magnitude of the estimated coefficients represent the relative importance of a particular terrain covariate. For example, a species with a large estimated coefficient on aspect is more likely to grow on a north facing slope. To identify species which will grow well on any given terrain, species with low-magnitude coefficients should be prioritized.

# Limitations

There is likely some large violation of the assumption of point-wise independence. From visual investigation of flower observations, there is evidence of some clustering, especially with respect to those species with limited observations. We recommend limiting inferences from this modeling approach to species with many observations that don't appear to violate the assumption of independence. Although not considered here, Ripley's K-function can quantify the degree of clustering for a given species. 

The assumption of a log-linear intensity with respect to terrain covariates limits the intensity to be exponential and uni-directional as a function of a given terrain covariate. This means that for a given species, the relationship between a particular terrain covariate and the estimated intensity can only increase or decrease in an exponential form. Considering that a species may have an "optimal" terrain type, future work should aim to fit a similar model that includes polynomial terrain coefficients, for example, $roughness^2$. The introduction of polynomial predictors will allow for bi-directional intensities as a function fo particular terrain covariates.

Additionally, interaction terms were not considered in these models, although the true existence of interaction effects may exist. For example slope may mediate the relationship of intensity with roughness. That is, a species may grow poorly on rough terrain with a low slope, but may grow particularly well on rough terrain with a high slope. Future modeling should consider the existance of interaction effects between terrain covariates. 

# Results

Given the sparsity of some the species, we limit inferences from the modeling approach to species with greater than 50 observations. 

```{r}
load("flower_df_final.RData")

print(flower_inf_50)
```


# Packages used

The R package `bookdown` [@R-bookdown] was used to create this report document using the `R` language [@R-base]. In addition, the following were packages used for the analysis and/or formatting of this document:

- `knitr` [@R-knitr]
- `rmarkdown` [@R-rmarkdown]
- `readr` [@R-readr]
- `readxl` [@R-readxl]
- `tibble` [@R-tibble]
- `dplyr` [@R-dplyr]
- `mosaic` [@R-mosaic]
- `nlme` [@R-nlme]
- `yarrr` [@R-yarrr]
- `ggplot2` [@R-ggplot2]
- `effects` [@R-effects]
- `predictmeans` [@R-predictmeans]

# References

